Parsing Postfix log files

### 2007-03-25 doc/logparser.tex
Lots of clarifications, improvements, more detail. Added statistics about (non-)compiled regexs and sorted/reversed/shuffled rule_order.

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible.  This paper describes the process required, documenting
    the algorithm and difficulties encountered, with reference to an
    implementation which stores data gathered from the logs in an SQL
    database.

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why mail is being rejected, or to check whether their new
anti-spam measures are working.  The more adventurous will have generated
some statistics to show how many hits each of their anti-spam measures have
gotten in the last week, and possibly even generated some graphs to clearly
illustrate the point to management or complaining users.\footnote{This was
the author's first real foray into processing Postfix logs.}  Very few will
have performed in-depth parsing and analysis of their logs, where the
parsing must correlate the log lines per-connection or per-queueid, rather
than processing lines independently.  One of the barriers to this type of
processing is the unstructured nature of Postfix's logs, where each logging
line was added on an ad hoc basis as a requirement was discovered or new
functionality was added.  Further complication arises the ability to have
rejection messages defined by the administrator, and every Real time Black
List returns a different message.  This paper documents the difficult
process of parsing Postfix logs, and presents a program which parses logs
and places the resulting data into a database for further processing by
other applications.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be applied and combined on a per-host,
per-recipient or per-sender basis, allowing clients/recipients to be
whitelisted with respect to specific restrictions.  Postfix leverages
simple lookup tables to support arbitrarily complicated user-defined
sequences of restrictions and exceptions, with the ultimate in flexibility
being the facility to consult an external process which can implement
whatever logic is required: if you wish you can restrict some users to
sending mail on the third Tuesday after pay day only, if that is of any use
to you.  Administrators can also supply their own rejection messages, to
make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the progress of an email through Postfix
can be quite difficult.  The logs tend to follow a 90\%/10\% pattern: 90\%
of the time following the log entries is simple, but the other 10\% of the
logging requires 90\% of the code.\footnote{These numbers don't have a
solid scientific basis, they're based on gut feeling from writing the
software.}

@@

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleaned from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet requirements, one can
implement one's own processing.

@@

Before getting into detail about the parser, a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the
tokeniser/Lex part of a Lex/YACC parser; they identify the line and return
some data, which the algorithm (the YACC part) receives, performing the
correct action.  Rules are all about identifying a line and extracting data
from it, whereas the algorithm is all about following the path the mail
takes through Postfix, piecing the data together into a coherent whole, and
performing housekeeping duties.

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without modifications to the
algorithm source code (significantly lowering the barrier to entry for new
or casual users), and greatly simplifies both algorithm and rules.  It also
allows for a clear separation of functionality: rules handle low level
details of identifying a line, whereas the algorithm handles the higher
level details of following the path a mail takes through postfix,
assembling the required data, etc.

Rule characteristics:

@@

    \item The first matching rule wins: no further rules are tried against
        that line,\footnote{With one exception, detailed later in Section
        \ref{actions-in-detail}.} but there is a facility for the specifying
        the order of the rules, so that more specific rules can be matched
        first.

    \item Rules are completely self-contained: there are no sub-rules, so
        rules have linear complexity, and can be understood without
        reference to any other rules.

@@

    \item [program] The program (smtp, smtpd, qmgr, etc) which the rule
        applies to.  This avoids needlessly trying rules which won't
        match the line, or worse, might match unintentionally.

    \item [regex] The regex to compare the log line against.  The regex
        will first have several keywords expanded; this simplifies reading
        and writing rules, avoids needless repetition of complex regex
        components, and allows the components to be corrected and/or
        improved in one location.

    \item [result\_cols, connection\_cols] This specifies how the fields in
        the log line will be extracted.  The format is: \newline
        client\_hostname = 1; helo = 2; sender = 4; \newline
        i.e.  semi-colon separated assignment statements, with the variable
        name on the left and the match from the regex (\$1, \$2 etc.) on the
        right hand side (without \$).

    \item [result\_data, connection\_data] Sometimes rules need to fake
        a piece of data which isn't present in the log line: the most
        common would be smtp\_code when mail is accepted.  The format of
        the same as result\_cols and connection\_cols.

    \item [result]  XXX RESULT NEEDS TO BE RENAMED XXX This is the action
        Postfix would have taken to generate this line.  Info is represents
        an unspecified intermediate action that the parser is not
        interested in per se, but which does log useful information.  The
        extracted information will be used, but the other results will be
        discarded.

@@

            \item [save\_by\_pid] Use the pid from the log line to lookup a
                connection, then save the data from the log line in the
                connection data structure.

            \item [commit] The mail is finished with, enter it in the
                database.

@@

    \item [rule\_order] This is an efficiency measure.  A counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in ascending rule\_order, and at the end of the run
        updates every rule's rule\_order so that rules which matched more
        frequently have a lower rule\_order.  Assuming that the 
        distribution of log lines is reasonably consistent, rules matching
        more commonly occurring log lines will be tried before rules
        matching less commonly occurring log lines, lowering the program's
        execution time.  XXX GIVE EXAMPLE TIMINGS XXX

@@

While the rules are more than double (treble with 2.3.x rules?) the size of
the parser, the rules are quite simple, and each rule is very much
independent of its fellows.  The parser is significantly more complicated,
and highly interdependent.

@@

    \item Follow the progress of the mail until it's either delivered or
        bounced, log all data, and finish.

@@

    \item The mail lacks a queueid until it has been accepted, so log lines
        must first be correlated by the smtpd pid, then transition to being
        correlated by the queueid.  This is relatively minor, but does
        require two versions of several functions, \texttt{by\_pid} and
        \texttt{by\_queueid}.

    \item Multiple independent mails may be delivered during one
        connection; this requires cloning the current data, so that
        subsequent mails won't trample over each other.  This must be done
        every time a mail is accepted, as it's impossible to tell in
        advance which connections will accept multiple mails.  It is quite
        easy to overlook this complication because only a small minority of
        connections accept more than one mail. Happily once the mail has
        been accepted log entries won't be correlated by pid for that mail
        any more, so there isn't any ambiguity about which mail a given log
        line belongs to.  The original connection will be discarded when
        the client disconnects.  One possible (as yet unseen) difficulty is
        distinguishing between rejects which belong to the current mail,
        and rejects which belong to previous, unaccepted, mails.

@@

        Unfortunately while all log lines from an individual process appear
        in the logs in monotonic order, the order in which log lines from
        different processes are interleaved is subject to the vagaries of
        process scheduling.  In addition the first log line belonging to
        the child mail (the log line cited above properly belongs to the
        parent mail) is logged by qmgr, so depending on how busy qmgr (and
        the server generally) is, the parent's creation log line may appear
        before or after (usually after) the child's log line from
        qmgr.\footnote{Postfix is quite paranoid about mail delivery, an
        excellent characteristic for an MTA to posess, so it won't log that
        the child has been created until it is absolutely certain that the
        mail has been written to disk.  Because the parent then has to be
        informed that the mail has been committed, it's most likely that
        the qmgr will log first.}

@@

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and removed
        \textbf{before} the parent logs the creation line!  Thus mails
        marked as faked cannot be committed, instead they must be marked as
        commit ready and subsequently committed by the parent mail.  Quite
        apart from identifying mail injected in an unknown fashion, or bugs
        in the parser, faked mails need to be marked as such because they
        lack some data present in the parent mail, and must copy that data
        from their parent.

>>>>

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible.  This paper describes the process required, documenting
    the parsing algorithm and rules, and explaining the difficulties
    encountered, with reference to an implementation which stores data
    gathered from the logs in an SQL database for later processing.

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated some statistics to show how many hits each of their anti-spam
measures has gotten in the last week, and possibly even generated some
graphs to clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid, rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix's logs, where each logging line was added on an ad hoc basis as
a requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator, and every Real-time Black List returns a
different explanatory message.  This paper documents the difficult process
of parsing Postfix logs, and presents a program which parses logs and
places the resulting data into a database for further processing by other
applications.

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be selectively applied, combined and
bypassed on a per-client, per-recipient or per-sender basis, allowing
varying levels of defense and or permissiveness.  Postfix leverages simple
lookup tables to support arbitrarily complicated user-defined sequences of
restrictions and exceptions, with the ultimate in flexibility being the
facility to consult an external process which can implement whatever logic
is required: some users can be restricted to sending mail on the third
Tuesday after pay day only, if that is of any use.\footnote{This example
may actually be useful in a payroll system.  More commonly encountered
scenarios are restricting posting to large mailing lists (e.g. sending
weekly offers to customers) to out of office hours, checking SPF records
(see http://www.openspf.org/ or
http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details), or
rate limiting clients.}  Administrators can also supply their own rejection
messages, to make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the journey an email takes through
Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
pattern: 90\% of the time the journey is simple, but the other 10\% of the
time requires 90\% of the code.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

@@

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are worth using in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The program should be reasonably easily
extensible, so if logging to a database doesn't meet requirements,
appropriate functionality can be implemented.

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser; rules identify the line and return some data,
which the algorithm (the YACC part) receives, performing the requested
action.  Rules are solely concerned with identifying a line and extracting
data from it, whereas the algorithm's task is to follow the journey each
mail takes through Postfix, piecing the data together into a coherent
whole, and performing housekeeping duties.

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without requiring modifications
to the algorithm source code (significantly lowering the barrier to entry
for new or casual users), and greatly simplifies both algorithm and rules.
It also allows for a clear separation of functionality: rules handle low
level details of identifying and extracting data from a line, whereas the
algorithm handles the higher level details of following the path a mail
takes through postfix, assembling the required data, etc.

Rule have certain characteristics which may help in understanding the
parser:

@@

    \item The first matching rule wins: no further rules are tried against
        that line,\footnote{With one exception, detailed later in Section
        \ref{actions-in-detail}.} but there is a facility for the
        specifying the order of the rules, so that more specific rules can
        be tried first.

    \item Rules are completely self-contained: there are no sub-rules, so
        rules have linear complexity, and can be understood in isolation,
        without reference to any other rules.

    \item XXX EXPAND ON THIS SOME MORE XXX

@@

    \item [program] The program (smtp, smtpd, qmgr, etc.) whose log lines
        the rule applies to.  This avoids needlessly trying rules which
        won't match the line, or worse, might match unintentionally.

    \item [regex] The regex to match the log line against.  The regex will
        first have several keywords expanded: this simplifies reading and
        writing rules, avoids needless repetition of complex regex
        components, and allows the components to be corrected and/or
        improved in one location.  For efficiency the keywords are expanded
        and every rule's regex is compiled before attempting to parse the
        log file - otherwise each regex would be recompiled each time it
        was used, resulting in an 468.95\% slowdown.

        Averaged over 10 test runs:

            \item Compiled run time was 55.47 seconds, standard deviation
                was 0.247 seconds.

            \item Non-compiled run time was 260.14 seconds (4 minutes,
                20.14 seconds), standard deviation was 0.96 seconds.

        Logging to the database was disabled for the test runs.

    \item [result\_cols, connection\_cols] Specifies how the fields in
        the log line will be extracted.  The format is: \newline
        client\_hostname = 1; recipient = 2; sender = 4; \newline
        i.e. semi-colon separated assignment statements, with the variable
        name on the left and the matching field from the regex on the
        right hand side.

    \item [result\_data, connection\_data] Sometimes rules need to fake a
        piece of data which isn't present in the log line: e.g. setting
        smtp\_code when mail is accepted.  The format is the same as
        result\_cols and connection\_cols, except that arbitrary data can
        be specified rather than just a number.

    \item [result]  XXX RESULT NEEDS TO BE RENAMED XXX This is the action
        Postfix must have taken to generate this line, with two exceptions:

            \item [info] Represents an unspecified intermediate action that
                the parser is not interested in per se, but which does log
                useful information, supplementing other log lines.

            \item [ignored] An action which is not only uninteresting in
                itself, but which also provides no useful data.

        Uninteresting lines are parsed rather than ignored completely so
        that any lines the parser isn't capable of handling become
        immediately obvious.

@@

            \item [save\_by\_pid] Use the pid from the log line to lookup a
                connection, then save the data from the log line in the
                connection data structure thus found.

            \item [commit] The mail is finished with, enter it in the
                database, and remove the mail from memory.

@@

    \item [rule\_order] This is an efficiency measure.  This counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in descending rule\_order, and at the end of the run
        updates every rule's rule\_order.  Assuming that the 
        distribution of log lines is reasonably consistent, rules matching
        more commonly occurring log lines will be tried before rules
        matching less commonly occurring log lines, lowering the program's
        execution time.

        Averaged over 10 test runs:

            \item Ascending order run time was 56.684 seconds, standard
                deviation was 0.29 seconds.

            \item Descending order (reversed) run time was 64.037 seconds,
                standard deviation was 0.16 seconds.

            \item Shuffled order run time was 57.19 seconds, standard
                deviation was 0.95 seconds.

        Logging to the database was disabled for the test runs.

@@

While the rules are more than double (treble with 2.3.x rules?) the size of
the algorithm, the rules are quite simple, and each rule is completely
independent of its fellows.  The parser is significantly more complicated,
and highly internally interdependent.

@@

    \item Follow the progress of the accepted mail until it's either delivered or
        bounced, log all data, and finish.

@@

    \item The mail lacks a queueid until it has been accepted, so log lines
        must first be correlated by the smtpd pid, then transition to being
        correlated by the queueid.  This is relatively minor, but does
        require:

            \item Two versions of several functions, \texttt{by\_pid} and
                \texttt{by\_queueid}.

            \item Two data structures to hold the data structures for each
                connection.

            \item Most importantly: every section of code must know whether
                it is needs to lookup the data structures by pid or
                queueid.

    \item Multiple independent mails may be delivered during one
        connection: this requires cloning the current data, so that
        subsequent mails won't trample over each other.  This must be done
        every time a mail is accepted, as it's impossible to tell in
        advance which connections will accept multiple mails.  It is quite
        easy to overlook this complication because only a small minority of
        connections accept more than one mail. Happily once the mail has
        been accepted log entries won't be correlated by pid for that mail
        any more (queueid will be used instead), so there isn't any
        ambiguity about which mail a given log line belongs to.  The
        original connection will be discarded unsaved when the client
        disconnects.  One unsolved difficulty is distinguishing between
        different groups of rejections, e.g.

            \item The client attempts to deliver a mail, but it is
                rejected.

            \item The client issues the RSET command to reset the session.

            \item The client attempts to deliver another mail, likewise
                rejected.

        There should probably be two different entires in the database
        resulting from the above sequence, but currently there will only be
        one.

@@

        Unfortunately while all log lines from an individual process appear
        in chronological order, the order in which log lines from different
        processes are interleaved is subject to the vagaries of process
        scheduling.  In addition the first log line belonging to the child
        mail (the log line cited above properly belongs to the parent mail)
        is logged by qmgr,\footnote{Qmgr is the Postfix daemon which
        manages the mail queue, determining which mails will have delivery
        attempted next.} so the order also depends on how busy qmgr
        is.\footnote{Postfix is quite paranoid about mail delivery, an
        excellent characteristic for an MTA to possess, so it won't log
        that the child has been created until it is absolutely certain that
        the mail has been written to disk.}

@@

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and removed
        \textbf{before} the parent logs the creation line!  Thus mails
        marked as faked cannot be entered into the database; instead they
        must be marked as database ready and subsequently entered by the
        parent mail.  Quite apart from identifying mail injected in an
        unknown fashion, or bugs in the parser, faked mails need to be
        marked as such because they lack some data present in the parent
        mail, and must copy that data from their parent.  XXX EITHER REMOVE
        THE LAST LINE OR EXPAND ON IT SOME MORE, AND FIGURE OUT WHY MORE
        DATA ISN'T REQUIRED XXX
